### 2020-04-06 Manuscript/manuscript.tex
small edits in response to reviewer line-by-line suggestions

\begin{figure}
\centering
\includegraphics[width=\textwidth]{../Figures/Tectonic_history.pdf}
\caption{\small{\textbf{Simplified timeline of Laurentia's tectonic history over the past $\sim$1.8 billion years.} Brief summaries and references related to the orogenic and rifting episodes are given in the text. A timeline of large igneous provinces  (LIPs) associated with typically brief and voluminous (or interpreted to be voluminous) volcanism is also shown. The interpreted age of paleomagnetic poles for Laurentia (not including separated terranes) compiled in this study for the Proterozoic and in \cite{Torsvik2012a} for the Phanerozoic is shown.}}
\label{fig:tectonic_history}
\end{figure}

@@

Following the amalgamation of the Archean provinces in Laurentia ca. 1.8 Ga, poles from each part of Laurentia can be considered to reflect the position of the entire composite craton. It is worth considering the possibility that poles from zones of Paleoproterozoic and Mesoproterozoic accretion could be allochthonous to the craton. \cite{Halls2015b} argued that this was the case for late Mesoproterozoic and early Neoproterozoic poles from east of the Grenvillian allochthon boundary fault. However, the majority of researchers have considered these poles to post-date major differential motion and be associated with cooling during collapse of a thick orogenic plateau developed during continent-continent collision (e.g. \citealp{Brown2012a}). Poles with a B-rating are also included in the composite that come from Greenland, Svalbard and Scotland. These terranes were once part of contiguous Laurentia, but have subsequently rifted away. These poles need to be rotated into the Laurentia reference frame prior to use for tectonic reconstruction, and I apply the rotations shown in Table \ref{tab:terrane_rotations}. The Euler pole and rotation is quite well-constrained for Greenland as it is associated with recent opening of Baffin Bay and the Labrador Sea (for which the rotation of \citealp{Roest1989a} is used). The reconstruction of Scotland is associated with the opening of the Atlantic (for which the rotation employed by \citealp{Torsvik2017a} is used) which is well-constrained but has more uncertainty associated with the Euler pole than that for Greenland. The reconstruction of Svalbard is more challenging given a multi-state tectonic history involving both translation within the Caledonides and subsequent rifting. The preferred Euler pole parameters of \cite{Maloof2006a} is used here. This Euler rotation is designed, in particular, to honor the high degree of similarity between Tonian sediments in East Greenland \citep{Hoffman2012a} and those of East Svalbard \citep{Maloof2006a} and to reconstruct East Svalbard to be aligned with these correlative sedimentary rocks.

@@

An exception to this overall agreement between poles from Greenland and mainland Laurentia occurs ca. 1382 Ma. There are poles of this age from Greenland associated with the Zig-Zag Dal basalts and related intrusions \citep{Marcussen1983a, Abrahamsen1987a}. However, these poles are in a distinct location from poles of similar age associated with the Belt Supergroup (e.g. the McNamara Formation and Pilcher/Garnet Range and Libby Formations; \citealp{Elston2002a}). Additionally, the older Belt Supergroup poles form a more southerly population than time-equivalent poles from elsewhere in Laurentia such as the Mistastin Pluton. There are potential complications associated with the Belt Supergroup being exposed within thrust sheets with significant Cenozoic Mesozoic and Cenozoic deformation. However, vertical axis rotations of the Belt region are not able to bring the Belt poles into agreement with those from Laurentia or Greenland nor is translation away from the craton. Another potential complication is that the remanence used for the development of the Belt Supergroup resides in hematite. As a result, there is the potential for inclination-flatting within the sedimentary rocks from which poles are developed. However, applying a moderate inclination factor of $f=0.6$ also does not bring the poles into congruence with the Zig-Zag basalts. There is the potential that the hematite could be the result of post-depositional oxidation (the remanence of the Purcell lavas pole is also held by hematite), however the overall coherency of the pole directions and the presence of reversals has been taken as evidence that the remanence is primary \citep{Elston2002a}. At present, it is unclear which poles are a better representation of Laurentia's position ca. 1400 Ma.

Another challenging portion of the Laurentia paleomagnetic record is that for the Ediacaran Period where there is little consistency between poles of similar age (Figs. \ref{fig:Laurentia_poles} and \ref{fig:Laurentia_paleolatitude}). As a result, there are poles that imply both low-latitude and high-latitude positions of Laurentia between 615 and 565 Ma. One explanation for these variable pole positions is that they are the result of large-scale oscillatory true polar wander in the Ediacaran that has influenced poles in Baltica and West Africa as well \citep{McCausland2007a, Robert2017a}. Another possibility is that the lack of congruency between poles in this point in the record is due to a particularly weak and non-dipolar geomagnetic field \citep{Abrajevitch2010a, Halls2015a, Bono2019a}. Regardless of mechanism, the Ediacaran data stand out as anomalous relative to the coherency of the rest of the poles in the composite (Fig. \ref{fig:Laurentia_paleolatitude}).

@@

A synthesized pole path for Laurentia is developed here and used to develop a continuous paleogeographic reconstruction of Laurentia constrained by the compilation of paleomagnetic poles. The paleolatitude implied by this continuous model is shown in Figure \ref{fig:Laurentia_paleolatitude}. This path is based on Laurentia data alone which means that it is poorly constrained through intervals of sparse data (950-850 Ma for example). One could use interpretations of paleogeographic connections with other cratons (e.g. Baltica in the early Neoproterozoic) to fill in such portions of the path, however the result then becomes model-dependent without being constrained by data from Laurentia itself. In portions of the record with a more dense record of poles, such as ca. 1450 Ma, a calculated running mean is used to integrate constraints from multiple poles. This method follows the approach taken in the Phanerozoic (e.g. \citealp{Torsvik2012a} wherein all poles within a 20 Myr interval are averaged with the interval than progressively moved forward in 10 Myr steps. When there are isolated `A' grade poles without other temporally-similar poles, these poles are fully satisfied in model. Where there are no constraints a simple interpolation between constraints is made. While data from Scotland and Svalbard are associated with Laurentia, the Scotland poles are poorly constrained in time and the Svalbard rotation to Laurentia is uncertain. These poles are not utilized in the simple Laurentia model which means that the model as shown does not include oscillatory true polar wander interpreted to have occurred ca. 810 and 790 Ma based on data from Svalbard \citep{Maloof2006a}. The model of \cite{Li2013a} shown in Figure \ref{fig:Laurentia_paleolatitude} does seek to incorporate this true polar wander while also incorporating an interpretation of the paleomagnetic pole record from South China.

@@

Paleogeographic snapshots for the past position of Laurentia reconstructed using this synthesis of the paleomagnetic poles are shown in Figure \ref{fig:Laurentia_reconstructions}. These reconstructions use the tectonic elements as defined by \citet{Whitmeyer2007a} with these elements being progressively added associated with Laurentia's accretionary growth. As a reminder to the reader, paleomagnetic poles provide constraints on the paleolatitude of a continental block as well as its orientation (which way was north relative to the block). While they provide constraints in this regard, they do not provide constraints in and of themselves for the longitudinal position of the block. Other approaches to obtain paleolongitude utilize geophysical hypotheses such as assuming that large low shear velocity provinces have been stable plume-generating zones in the lower mantle to which plumes can be reconstructed \citep{Torsvik2014a} or that significant pole motion in certain time intervals is associated with true polar wander axes that switch through time in conjunction with the supercontinent cycle \citep{Mitchell2012a}. In Figure \ref{fig:Laurentia_reconstructions}, Laurentia is centered on the longitudinal position of Duluth with the orientation and paleolatitude being constrained by the paleomagnetic pole compilation as synthesized in the simple Laurentia pole interpolation model. The continuous paleolatitude implied by this model is shown as the simple Laurentia pole interpolation mode in Figure \ref{fig:Laurentia_paleolatitude}.

@@

Another perspective on Proterozoic tectonics, is that the record is one of intermittent subduction \citep{Silver2008a, ONeill2013a}. In such a model, there are extended intervals with a stagnant lid alternating with intervals of differential plate motion. In particular, it has been argued that the Mesoproterozoic Era (1.6 to 1.0 Ga) is an interval when Earth was in a stagnant regime \citep{Silver2008a, ONeill2013a}. The accretionary history of Laurentia following the amalgamation of the Archean provinces is difficult to reconcile with such an interpretation (Figs. \ref{fig:Laurentia_map} and \ref{fig:tectonic_history}). An additional constraint comes from the paleomagnetic record and paleomagnetic poles supported with baked contact tests through the Mesoproterozoic. In a stagnant lid regime, there would not be sufficient heat flow across the core-mantle boundary to sustain a geodynamo \citep{Nimmo2000a, Buffett2000b}. Baked contact tests indicate that, at the time of dike emplacement, there was an appreciable field such that both the cooling magma and the heated country rock in the vicinity of a dike were able to acquire a primary coherent magnetization direction. Additionally, since paleomagnetic poles are developed from many individual cooling units across a region, the similarity of the directions across an igneous province indicates that the magnetizations were dominantly acquired from the geomagnetic field rather than being influenced by local variable crustal magnetizations.  Therefore, the record supports the persistence of a geomagnetic field through the Paleoproterozoic and Mesoproterozoic (Table 2) which implies active plate tectonics that enabled sufficient core-mantle boundary heat flow to power the geodynamo.  This interpretation of a significant persistent geomagnetic field through much of the Proterozoic (with the potential exception of the Ediacaran; \citealp{Bono2019a}) is further bolstered by estimates of paleointensity obtained from mafic dikes from Laurentia (e.g. \citealp{Macouin2006a}) and elsewhere.

>>>>

\begin{figure}
\centering
\includegraphics[width=\textwidth]{../Figures/Tectonic_history.pdf}
\caption{\small{\textbf{Simplified timeline of Laurentia's tectonic history over the past $\sim$1.8 billion years.} Brief summaries and references related to the orogenic and rifting episodes are given in the text. A timeline of large igneous provinces (LIPs) associated with typically brief and voluminous (or interpreted to be voluminous) volcanism is also shown. The interpreted age of paleomagnetic poles for Laurentia (not including separated terranes) compiled in this study for the Proterozoic and in \cite{Torsvik2012a} for the Phanerozoic is shown. Abbreviations on the figure: CAMP (Central Atlantic Magmatic Province); proto-Cord (proto-Cordilleran).}}
\label{fig:tectonic_history}
\end{figure}

@@

Following the amalgamation of the Archean provinces in Laurentia ca. 1.8 Ga, poles from each part of Laurentia can be considered to reflect the position of the entire composite craton. It is worth considering the possibility that poles from zones of Paleoproterozoic and Mesoproterozoic accretion could be allochthonous to the craton. \cite{Halls2015b} argued that this was the case for late Mesoproterozoic and early Neoproterozoic poles from east of the Grenvillian allochthon boundary fault. However, the majority of researchers have considered these poles to post-date major differential motion and be associated with cooling during collapse of a thick orogenic plateau developed during continent-continent collision (e.g. \citealp{Brown2012a}). Poles with a B-rating are also included in the compilation that come from Greenland, Svalbard and Scotland. These terranes were once part of contiguous Laurentia, but have subsequently rifted away. These poles need to be rotated into the Laurentia reference frame prior to use for tectonic reconstruction, and I apply the rotations shown in Table \ref{tab:terrane_rotations}. The Euler pole and rotation is quite well-constrained for Greenland as it is associated with recent opening of Baffin Bay and the Labrador Sea (for which the rotation of \citealp{Roest1989a} is used). The reconstruction of Scotland is associated with the opening of the Atlantic (for which the rotation employed by \citealp{Torsvik2017a} is used) which is well-constrained but has more uncertainty associated with the Euler pole than that for Greenland. The reconstruction of Svalbard is more challenging given a multi-state tectonic history involving both translation within the Caledonides and subsequent rifting. The preferred Euler pole parameters of \cite{Maloof2006a} are used here for this reconstruction. This Euler rotation is designed, in particular, to honor the high degree of similarity between Tonian sediments in East Greenland \citep{Hoffman2012a} and those of East Svalbard \citep{Maloof2006a} and to reconstruct East Svalbard to be aligned with these correlative sedimentary rocks.

@@

An exception to this overall agreement between poles from Greenland and mainland Laurentia occurs ca. 1382 Ma. There are poles of this age from Greenland associated with the Zig-Zag Dal basalts and related intrusions \citep{Marcussen1983a, Abrahamsen1987a}. However, these poles are in a distinct location from poles of similar age associated with the Belt Supergroup (e.g. the McNamara Formation and Pilcher/Garnet Range and Libby Formations; \citealp{Elston2002a}). Additionally, the older Belt Supergroup poles form a more southerly population than time-equivalent poles from elsewhere in Laurentia such as the Mistastin Pluton. There are potential complications associated with the Belt Supergroup being exposed within thrust sheets with significant Cenozoic Mesozoic and Cenozoic deformation. However, vertical axis rotations of the Belt region are not able to bring the Belt poles into agreement with those from Laurentia or Greenland nor is translation away from the craton. Another potential complication is that the remanence used for the development of the Belt Supergroup resides in hematite. As a result, there is the potential for inclination-flatting within the sedimentary rocks from which poles are developed. However, applying a moderate inclination factor of $f=0.6$ also does not bring the poles into congruence with the Zig-Zag basalts. There is the potential that the hematite could be the result of post-depositional oxidation (the remanence of the lavas pole is also held by hematite), however the overall coherency of the pole directions and the presence of reversals has been taken as evidence that the remanence is primary \citep{Elston2002a}. At present, it is unclear which poles are a better representation of Laurentia's position ca. 1400 Ma.

Another challenging portion of the Laurentia paleomagnetic record is that for the Ediacaran Period where there is little consistency between poles of similar age (Figs. \ref{fig:Laurentia_poles} and \ref{fig:Laurentia_paleolatitude}). As a result, there are poles that imply both low-latitude and high-latitude positions of Laurentia between 615 and 565 Ma. One explanation for these variable pole positions is that they are the result of large-scale oscillatory true polar wander in the Ediacaran where rapid rotation of the entire silicate Earth influenced poles in Baltica and West Africa as well \citep{McCausland2007a, Robert2017a}. Another possibility is that the lack of congruency between poles in this point in the record is due to a particularly weak and non-dipolar geomagnetic field \citep{Abrajevitch2010a, Halls2015a, Bono2019a}. Regardless of mechanism, the Ediacaran data stand out as anomalous relative to the coherency of the rest of the poles in the composite (Fig. \ref{fig:Laurentia_paleolatitude}).

@@

A synthesized pole path for Laurentia is developed here and used to develop a paleogeographic reconstruction of Laurentia constrained by the compilation of paleomagnetic poles. The paleolatitude implied by this continuous model is shown in Figure \ref{fig:Laurentia_paleolatitude}. This path is based on Laurentia data alone which means that it is poorly constrained through intervals of sparse data (950-850 Ma for example). One could use interpretations of paleogeographic connections with other cratons (e.g. Baltica in the early Neoproterozoic) to fill in such portions of the path, however the result then becomes model-dependent without being constrained by data from Laurentia itself. In portions of the record with a more dense record of poles, such as ca. 1450 Ma, a calculated running mean is used to integrate constraints from multiple poles. This method follows the approach taken in the Phanerozoic (e.g. \citealp{Torsvik2012a} wherein all poles within a 20 Myr interval are averaged with the interval than progressively moved forward in 10 Myr steps. When there are isolated `A' grade poles without other temporally-similar poles, these poles are fully satisfied in model. Where there are no constraints a simple interpolation between constraints is made. While data from Scotland and Svalbard are associated with Laurentia, the Scotland poles are poorly constrained in time and the Svalbard rotation to Laurentia is uncertain. These poles are not utilized in the simple Laurentia model which means that the model as shown does not include oscillatory true polar wander interpreted to have occurred ca. 810 and 790 Ma based on data from Svalbard \citep{Maloof2006a}. The model of \cite{Li2013a} shown in Figure \ref{fig:Laurentia_paleolatitude} does seek to incorporate this true polar wander while also incorporating an interpretation of the paleomagnetic pole record from South China.

@@

Paleogeographic snapshots for the past position of Laurentia reconstructed using this synthesis of the paleomagnetic poles are shown in Figure \ref{fig:Laurentia_reconstructions}. These reconstructions use the tectonic elements as defined by \citet{Whitmeyer2007a} with these elements being progressively added associated with Laurentia's accretionary growth. As a reminder to the reader, paleomagnetic poles provide constraints on the paleolatitude of a continental block as well as its orientation (which way was north relative to the block). While they provide constraints in this regard, they do not provide constraints in and of themselves for the longitudinal position of the block. Other approaches to obtain paleolongitude utilize geophysical hypotheses such as assuming that large low shear velocity provinces have been stable plume-generating zones in the lower mantle to which plumes can be reconstructed \citep{Torsvik2014a} or that significant pole motion in certain time intervals is associated with true polar wander axes that switch through time in conjunction with the supercontinent cycle \citep{Mitchell2012a}. In Figure \ref{fig:Laurentia_reconstructions}, Laurentia is centered on the longitudinal position of Duluth with the orientation and paleolatitude being constrained by the paleomagnetic pole compilation as synthesized in the simple Laurentia pole interpolation model (Fig. \ref{fig:Laurentia_paleolatitude}).

@@

Another perspective on Proterozoic tectonics, is that the record is one of intermittent subduction \citep{Silver2008a, ONeill2013a}. In such a model, there are extended intervals with a stagnant lid alternating with intervals of differential plate motion. In particular, it has been argued that the Mesoproterozoic Era (1.6 to 1.0 Ga) is an interval when Earth was in a stagnant regime \citep{Silver2008a, ONeill2013a}. The long-lived accretionary history of Laurentia following the amalgamation of the Archean provinces is difficult to reconcile with such an interpretation (Figs. \ref{fig:Laurentia_map} and \ref{fig:tectonic_history}).

An additional constraint supporting ongoing plate tectonics throughout the Proterozoic comes from the paleomagnetic record --- in particular the paleomagnetic poles supported with baked contact tests (Fig. \ref{fig:baked_contact}). In a stagnant lid regime, there would not be sufficient heat flow across the core-mantle boundary to sustain a geodynamo \citep{Nimmo2000a, Buffett2000b}. Baked contact tests indicate that, at the time of dike emplacement, there was an appreciable field such that both the cooling magma and the heated country rock in the vicinity of a dike were able to acquire a primary coherent magnetization direction. Additionally, since paleomagnetic poles are developed from many individual cooling units across a region, the similarity of the directions across an igneous province indicates that the magnetizations were dominantly acquired from the geomagnetic field rather than being influenced by local variable crustal magnetizations.  Therefore, the record supports the persistence of a geomagnetic field through the Paleoproterozoic and Mesoproterozoic (Table 2) which implies active plate tectonics that enabled sufficient core-mantle boundary heat flow to power the geodynamo.  This interpretation of a significant persistent geomagnetic field through much of the Proterozoic (with the potential exception of the Ediacaran; \citealp{Bono2019a}) is further bolstered by estimates of paleointensity obtained from mafic dikes from Laurentia (e.g. \citealp{Macouin2006a}) and elsewhere.
